---
title: "Selection of peptides for retention time tracking"
author: "Ozan Aygun"
date: "May 2017"
output:
  html_document:
    depth: 3
    highlight: tango
    number_sections: yes
    theme: cerulean
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE, fig.align='center',fig.width=5,fig.height=4}
knitr::opts_chunk$set(echo =FALSE, include = TRUE, message = FALSE,warning = FALSE)
```

# Introduction

# Choosing the relevant QC directories for peptide selection

In order to generate the desired peptide list, we selected the following JurkatQC runs:

```{r}
#################################################################
#
# Selection of JurkatQC directories to generate peptide reports
#
#################################################################


setwd("Z:/LabMembers/Ozan/RetentionTimePeptides")
# Get the updated QM list

QM <- readRDS("C:/Users/ozan/Desktop/Ozan_R_utilities/Development/JurkatQCscraper/QMreportArchive.rds")

QM_subset <- QM[which(QM$DistinctPepsCSTotal.number_of. > 28000),]

QM_Franklin <- QM_subset[QM_subset$instrument == "Franklin",][1:3,]
QM_Hubble2 <- QM_subset[QM_subset$instrument == "Hubble2",][1:3,]
QM_Galileo <- QM_subset[QM_subset$instrument == "Galileo",][1:3,]


QM_Beaker <-  QM_subset[(QM_subset$instrument == "Beaker")&(QM_subset$DistinctPepsCSTotal.number_of. >40000),][1,]
QM_Yoda <- QM_subset[(QM_subset$instrument == "Yoda")&(QM_subset$DistinctPepsCSTotal.number_of. >40000),][1,]


QM.summary <- rbind(QM_Galileo,QM_Hubble2,QM_Franklin, QM_Yoda,QM_Beaker)

QM.summary <- dplyr::select(QM.summary, -FileName_instrument_match,
                            -foo,-time,-date)

QM.summary.brief <- dplyr::select(QM.summary, File,MedianMS1peakwidthmid90.matchedspectra.sec.,DistinctPepsCSTotal.number_of.,MedianMS2filltimemid90.matchedspectra.msec. )
knitr::kable(QM.summary, align = "c", format = "pandoc",row.names = F)
```

In addition we included 2 HeLa Thermo Standard Runs that are previously searched at:

- bennett:/msdataSM/Lauren/InstrumentEvaluation/LumosEvaluation_HeLaCurve
- bennett:/msdataSM/Lauren/Hubble2_HeLaCurve

Spectrum Mill protein-peptide report was prepared this file was obtained:

- coston:/JurkatQC/Galileo/2017Q2/20170501/proteinPeptideComparisonColumnsExport.S.1.ssv


```{r, cache=TRUE}

setwd("Z:/LabMembers/Ozan/RetentionTimePeptides")

# Initial loading:
 pPeptide <- read.csv2("T:/JurkatQC/Galileo/2017Q2/20170501/proteinPeptideComparisonColumnsExport.S.1.ssv",stringsAsFactors = FALSE, na.strings = c(""))

 # Save as RDS for easy loading in future sessions
 saveRDS(pPeptide,file = "pPeptide.rds")
 
setwd("Z:/LabMembers/Ozan/RetentionTimePeptides")
pPeptide <- readRDS("pPeptide.rds")
repeating.names <- names(pPeptide)[1:16]
unique.names <- names(pPeptide)[209:233]

# Better to adjust the header:

first.row <- make.names(pPeptide[1,])
names(pPeptide) <- paste(names(pPeptide),first.row,sep = "__")
pPeptide <- pPeptide[-1,]

# Which sequences are identified in all QC runs?
w <- which(grepl("^Sequence|^sequence__",names(pPeptide)))

peptide.sequences <- pPeptide[,w]
comp.Peptide <- pPeptide[complete.cases(pPeptide[,w]),]
```

In this report, we obtained **`r length(unique(pPeptide$sequence__NA.)) `** unique peptides. **`r length(unique(comp.Peptide$sequence__NA.)) `** of these peptides were identified in all QC runs.

```{r}
# collect the relevant features from the complete peptide set

relevant.features <- c("^bestScore", "^totalIntensity", "^retentionTimeMin","^accession_number", "^best_parent_m_over_z", "^sequence__NA.")

w <- unlist(sapply(relevant.features, function(y){
        which(grepl(y,names(comp.Peptide) ))
}))

relevant.pep <- comp.Peptide[,w]


names(relevant.pep)[grepl("^bestScore",names(relevant.pep))] <- "bestScore"

relevant.pep$bestScore <- as.numeric(relevant.pep$bestScore)

# Define a score cut off

  
```

# Choosing a 'score cut off'

First we choose a bestScore cutoff:

```{r}
library(ggplot2)

qplot(x = bestScore,data = relevant.pep,geom = "histogram", bins = 100)+theme_bw()
  
```

Scores have a near Gaussian distribution with a mean `r round(mean(relevant.pep$bestScore),2)`. We choose to include peptides that have the bestScores above this mean score.

```{r}
library(dplyr)
relevant.pep <- dplyr::filter(relevant.pep, bestScore > mean(bestScore))

```

This reduced the peptide list to **`r length(relevant.pep$bestScore)`** unique peptides.

# Distribution of retention time



```{r, fig.width=9}

# First simplify the data

total.intensity <- relevant.pep[,which(grepl("totalIntensity",names(relevant.pep)))]
retentionTimeMin <- relevant.pep[,which(grepl("retentionTimeMin",names(relevant.pep)))]

total.intensity <- apply(total.intensity,2,as.numeric)
retentionTimeMin <- apply(retentionTimeMin,2,as.numeric)

relevant.pep$median.total.intensity <- apply(total.intensity,1,median)
relevant.pep$median.retentionTimeMin <-apply(retentionTimeMin,1,median)

hist(relevant.pep$median.retentionTimeMin, breaks = 20, col = "lightgreen",
     main = "Distribution of Median Retention Time (Min)",
     xlab = "Median Retention Time Across all QC runs")
```

```{r}
# Simplified data
simp.peptide <- dplyr::select(relevant.pep,sequence__NA., bestScore, median.retentionTimeMin,
                              median.total.intensity,best_parent_m_over_z__NA.,
                              accession_number__NA.,accession_numbers__NA.)

w <- which(grepl("__NA.$",names(simp.peptide)))
names(simp.peptide)[w] <- gsub("__NA.$","",names(simp.peptide)[w])
simp.peptide$best_parent_m_over_z <- as.numeric(simp.peptide$best_parent_m_over_z)
```

# Distribution of m/z

This subset of peptides have near gaussian m/z distribution:

```{r}
hist(simp.peptide$best_parent_m_over_z,col = "blue")
```

# Make 8 retention time bins

We will partition the data based on 8 retention time bins as we discussed in our meeting:

```{r}
#percentiles = quantile(simp.peptide$median.retentionTimeMin,probs = seq(0,1,0.1))
```



```{r,fig.width=12, fig.align="right", fig.height=4}
simp.peptide$median.retentionTime.Bins <-cut(simp.peptide$median.retentionTimeMin,8)

ggplot(aes(x = best_parent_m_over_z),data = simp.peptide )+
        geom_histogram(fill = "blue", color = "black")+ facet_grid(. ~ median.retentionTime.Bins)+
        theme_bw()+
        ggtitle("Retention time bins (8 x 11 min bins)")
```

- We notice that the first 2 retention time bins (up to 35 minutes), we have limited representation of m/z, these domain mainly consists of peptides with m/z that are below the average m/z.
- The rest of the retention time looks like a good representation of the overall m/z distribution. 



```{r,fig.width=12, fig.align="right", fig.height=4}
simp.peptide$median.retentionTime.Bins <-cut(simp.peptide$median.retentionTimeMin,8)

ggplot(aes(x = best_parent_m_over_z, y = log(median.total.intensity)),data = simp.peptide )+
        geom_point()+ facet_grid(. ~ median.retentionTime.Bins)+
        theme_bw()+
        ggtitle("Retention time bins (8 x 11 min bins)")
```


# Algorihtm for the selection of peptides

Peptides will be selected based on the following algorithm:

- Divide the retention time span into 8 percentile bins.
- For each retention time bin;
- Rank peptides based on total intensity